---
title: "Assignment 5: Data Visualization"
author: "Student Name"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A05_DataVisualization.Rmd") prior to submission.

The completed exercise is due on Monday, February 14 at 7:00 pm.



## Set up your session 
1. Set up your session. Verify your working directory and load the tidyverse 
and cowplot packages. Upload the NTL-LTER processed data files for nutrients 
and chemistry/physics for Peter and Paul Lakes (use the tidy 
[`NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv`] version) and 
the processed data file for the Niwot Ridge litter dataset (use the [`NEON_NIWO_Litter_mass_trap_Processed.csv`] version). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r}
#1 
getwd()
library(tidyverse)
library(cowplot)

NTL_LTER<-read.csv("./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv", stringsAsFactors = TRUE)

Niwot_Ridge<-read.csv("./Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv", stringsAsFactors = TRUE)

#2 
NTL_LTER$sampledate<-as.Date(NTL_LTER$sampledate, format = "%Y-%m-%d")

Niwot_Ridge$collectDate<-as.Date(Niwot_Ridge$collectDate, format = "%Y-%m-%d")

class(Niwot_Ridge$collectDate) 
class(NTL_LTER$sampledate)#now they're dates!
```
## Define your theme
3. Build a theme and set it as your default theme.

```{r}
#3
mytheme <- theme_classic(base_size = 14)+
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and `ylim()`).

```{r}
#4
tp_ug_vs_po4<-
  ggplot(NTL_LTER, aes(x = tp_ug, y = po4))+
  geom_point()+
  facet_wrap(vars(lakename))+
  xlim(0,140)+
  ylim(0,50)+
  geom_smooth(method = lm, color = "black")
print(tp_ug_vs_po4)

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned.

```{r}
#5
NTL_LTER$month<-as.factor(NTL_LTER$month) #forcing "month" to be a factor instead of a numeric variable so I can see multiple box plots

temp_boxplot<-
  ggplot(NTL_LTER, aes(x = month, y = temperature_C))+
  geom_boxplot(aes(color = lakename))
print(temp_boxplot)

tp_boxplot<-
  ggplot(NTL_LTER, aes(x = month, y = tp_ug))+
  geom_boxplot(aes(color = lakename))
print(tp_boxplot)

tn_boxplot<-
  ggplot(NTL_LTER, aes(x = month, y = tn_ug))+
  geom_boxplot(aes(color = lakename))
print(tn_boxplot)

#cowplot:
plot_grid(temp_boxplot,tp_boxplot, tn_boxplot, nrow = 1)

```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r}
#6

#7


```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer:
